A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-09-07 masters_thesis/si_paper/paper.tex
More modifications from reviewers; almost done

The ideal topology and swarm size for PSO depend on the objective function.
Researchers have devised various benchmark functions and have found that the
ideal topology for one function may perform very poorly for another function.
The No Free Lunch Theorems for Optimization show that this is true in
general---if an algorithm performs well on average for one class of functions
then it must do poorly on average for other
problems~\citep{wolpert-1997-nfl-for-optimization}.

An attempt to standardize PSO found that although a Complete swarm of 50
particles converged to the global optimum more quickly for many benchmark
functions than the same swarm size with a Ring topology, it was also more
likely to prematurely converge to local optima for other functions.  The study
found no significant improvement for any other swarm size between 20 and 100
and concluded with a recommendation to use a swarm of 50 particles with a Ring
topology as a starting point.  The authors acknowledged that choosing the ideal
topology requires thorough experimentation for the particular
problem~\citep{bratton-2007-defining-a-standard-for-pso}.  Other authors have
explained that a large swarm with a sparse topology propagates information
slowly~\citep{montes-de-oca-2009-frankensteins-pso}.

@@

There are many ways to parallelize the basic PSO algorithm.  The most
fundamental decision to make in parallel PSO is which parallel architecture to
use.  Several architectures have been proposed, including Master-Slave, fully
distributed (sometimes called ``diffusion''), and reformulating PSO into
Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
mcnabb-2007-parallel-pso-using-mapreduce}.

@@

Rastrigin is a multi-modal function that is best optimized with a large,
Complete swarm.  It has been shown that with Rastrigin, the more particles
there are in the swarm, the lower function value it finds, up to at least 4000
particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima and converge to poorer values.  Because our speculative
algorithms require significantly smaller swarm sizes, we would expect to not
perform very well on functions such as Rastrigin.  Our experiments show our
intuition to be correct.  In this experiment we used 240 processors, so SEPSO
and Pick Best each had 30 particles, and the Standard algorithms had 240
particles.  As expected, SEPSO and Pick Best converge quickly to worse local
optima than Standard does.  \fig{rastrigin} shows the results graphically, and
\tabref{rastrigin-20} shows results for all of the methods we tried.

@@

Given our observation from the 20 dimensional benchmark functions that Many
Iterations most often outperforms other speculative approaches, we only present
results for Many Iterations and Standard for the 50 and 500 dimensional
variants.  For these experiments we used 800 processors, as the problems are
more difficult and speculative approaches perform best when many processors are
available, as we showed in \sec{griewank-20}.  Many Iterations Complete showed
the same premature convergence that was observed with the 20 dimensional
benchmark functions, so we do not include those results in our tables; the
success rate was 0\% for all functions.

@@

Large parallel clusters are often required to successfully optimize practical
modern problems.  To properly use PSO with such clusters, a balance needs to be
made between using processors to increase the swarm size and using them to
increase the speed of the algorithm.  This work is a first step in that
direction that opens the door to many future improvements on speculative
methods in the parallelization of PSO.

@@

\section{Implementing Speculative Evaluation}
\label{sec:implementation}

@@

\subsection{Terminology}

@@

\subsection{Centralized Algorithms}

@@

\subsection{Distributed Algorithms}

@@

\subsection{Dynamic Topologies}

@@

\section{Alternate form of message passing}

>>>>

The ideal topology and swarm size for PSO depend on the objective function.
Researchers have devised various benchmark functions and have found that the
ideal topology for one function may perform very poorly for another function.
The No Free Lunch Theorems for Optimization show that this is true in
general---if an algorithm performs well on average for one class of functions
then it must do poorly on average for other
problems~\citep{wolpert-1997-nfl-for-optimization}.  A good starting point
seems to be a swarm of 50 particles using a Ring
topology~\citep{bratton-2007-defining-a-standard-for-pso}.

@@

There are many ways to parallelize the basic PSO algorithm.  The most
fundamental decision to make in parallel PSO is which parallel architecture to
use.  Several architectures have been proposed, including Master-Slave, fully
distributed (sometimes called ``diffusion''), and reformulating PSO into
Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
mcnabb-2007-parallel-pso-using-mapreduce}.  For inexpensive objective
functions, graphics processors (GPUs) have also been used to parallelize PSO
using either texture rendering~\citep{li-2007-fine-grained-parallel-pso-gpu} or
a general purpose GPU API~\citep{zhou-2009-gpu-based-parallel-pso}.

@@

Rastrigin is a multi-modal function that is best optimized with a large,
Complete swarm.  It has been shown that with Rastrigin, the more particles
there are in the swarm, the lower function value it finds, up to at least 4000
particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima and converge to poorer values.  In fact, none of the algorithms
that we tested were able to solve this problem optimally.  We thus set a
somewhat arbitrary success threshold, simply to be able to compare algorithms.

Because our speculative algorithms require significantly smaller swarm sizes,
we would expect to not perform very well on functions such as Rastrigin.  Our
experiments show our intuition to be correct.  In this experiment we used 240
processors, so SEPSO and Pick Best each had 30 particles, and the Standard
algorithms had 240 particles.  As expected, SEPSO and Pick Best converge
quickly to worse local optima than Standard does.  \fig{rastrigin} shows the
results graphically, and \tabref{rastrigin-20} shows results for all of the
methods we tried.

@@

Given our observation from the 20 dimensional benchmark functions that Many
Iterations most often outperforms other speculative approaches, we only present
results for Many Iterations and Standard for the 50 and 500 dimensional
variants.  We used 800 processors for these experiments because the problems
are more difficult, and because speculative approaches perform best when many
processors are available, as we showed in \sec{griewank-20}.  Many Iterations
Complete showed the same premature convergence that was observed with the 20
dimensional benchmark functions, so we do not include those results in our
tables; the success rate was 0\% for all functions.

@@

Large parallel clusters are often required to successfully optimize practical
modern problems, and the world's computing resources are increasingly coming
from such clusters.  To properly use PSO with these systems, a balance needs to
be made between using processors to increase the swarm size and using them to
increase the speed of the algorithm.  This work is a first step in that
direction that opens the door to many future improvements on speculative
methods in the parallelization of PSO.

@@

\section*{Appendix A Implementing Speculative Evaluation}
\label{sec:implementation}

@@

\subsection*{A.1 Terminology}

@@

\subsection*{A.2 Centralized Algorithms}

@@

\subsection*{A.3 Distributed Algorithms}

@@

\subsection*{A.4 Dynamic Topologies}

@@

\section*{Appendix B Alternate form of message passing}
